---
title: "Stat459 HW2"
author: "Zhixi Tan"
date: "September 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=5, fig.width=6.5)
```

```{r}
library(survival)
library(asaur)
```
##Read the data into data frame
```{r}
#(a)
#read the data into data frame
diabetes.dat <- read.csv("diabetes.csv")

diabetes.dat$sexF <- factor(diabetes.dat$sex, labels=c('M','F'))

#summary of the dataframe
summary(diabetes.dat)

```
##KM survival curves for two types of diab

```{r}
#(b)
#KM survival curves for two types of diab
plot(survfit(Surv(diabetes.dat$stime, diabetes.dat$deathind) ~ diabetes.dat$diab), col = 1:2, ylab = 'Survival prob', xlab = 'time in days', lty = 1:2)
title('KM curves, two types of diab')
legend('topright', c('Not Diabetic', 'Diabetic'), col = 1:2, lty = 1:2, bty = 'n')

```

##Survival probabilities for each diab type at one and five years using KM formula and FH formula
```{r}
#(c)

#convert the days into years
stimeYears <- diabetes.dat$stime / 365

#Survival prob. for each type at one and five years using Kaplan-Meier formula
md.km <- survfit(Surv(stimeYears, diabetes.dat$deathind) ~ diabetes.dat$diab, conf.type = "none")

summary(md.km, c(1,5))


#Survival prob. for each type at one and five years using Fleming-Harrington formula
md.fh <- survfit(Surv(stimeYears, diabetes.dat$deathind) ~ diabetes.dat$diab, type='fh', conf.type = "none")

summary(md.fh, c(1,5))
```

##Confidence interval using Kaplan-Meier formula
```{r}
#(d)

# Confidence interval using Kaplan-Meier formula with plain type
md.km2 <- survfit(Surv(stimeYears, diabetes.dat$deathind) ~ diabetes.dat$diab,
                 conf.type = "plain")
summary(md.km2, c(1,5))


#  Confidence interval using Kaplan-Meier formula with log-log type
md.km3 <- survfit(Surv(stimeYears, diabetes.dat$deathind) ~ diabetes.dat$diab,
                 conf.type = "log-log")
summary(md.km3, c(1,5))

```
#Confidence interval using Fleming-Harrington formula
```{r}
# Confidence interval using Fleming-Harrington formula with plain type
md.fh2 <- survfit(Surv(stimeYears, diabetes.dat$deathind) ~ diabetes.dat$diab, type='fh',
                conf.type = "plain" )
summary(md.fh2, c(1,5))


# Confidence interval using Fleming-Harrington formula with log-log type
md.fh3 <- survfit(Surv(stimeYears, diabetes.dat$deathind) ~ diabetes.dat$diab, type='fh',
                conf.type = "log-log" )
summary(md.fh3, c(1,5))

```
##Median survival time in years and confidence interval (using Kaplan-Meier formula with log-log type)
```{r}
#(e) Median survival time and confidence interval (using Kaplan-Meier formula with log-log type)

md.km3

```
##Log-rank test for equality of survival curves for two types of diab
```{r}
#(f)

#log-rank test for equality of survival curves for two types of diab
survdiff(Surv(diabetes.dat$stime, diabetes.dat$deathind) ~ diabetes.dat$diab)
```

##Median follow-up time
```{r}
#(g)

#median follow-up time

#First method: reverse KM
delta.followup <- 1 - diabetes.dat$deathind
survfit(Surv(diabetes.dat$stime, delta.followup) ~ 1)


#Second method: just median for the all of the survival time
median(diabetes.dat$stime)
```
##Cumulative hazard functions for two types of diab
```{r}
#(h)

#Cumulative hazard functions for two types of diab

plot(survfit(Surv(diabetes.dat$stime, diabetes.dat$deathind) ~ diabetes.dat$diab), 
     conf.int = T, mark="|", xlab = "Time in days",
     fun="cumhaz", ylab="Cumulative hazard", col = 1:2, lty = 1:2)
title("Cumulative hazard")
legend('topright', c('Not Diabetic', 'Diabetic'), col = 1:2, lty = 1:2, bty = 'n')
```

##Log-rank test for the equality of survival curves for two types of diab, but also stratify by sex
```{r}
#(i)

#log-rank test for the equality of survival curves for two types of diab, but also stratify by sex
survdiff(Surv(diabetes.dat$stime, diabetes.dat$deathind) ~ diabetes.dat$diab + strata(diabetes.dat$sexF))

```

