---
title: "Chapter 5"
author: "Mary Lesperance"
date: "September 24, 2018"
geometry:  margin=2in
output:
  html_document:
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=5, fig.width=6.5)
```

```{r}
# install.packages("asaur")  # this must be done once
library(survival)
library(asaur)
library(numDeriv)
library(rms)
library(tableone)
```

# Section 5.2, simple example, data from Section 4.1

## Figure 5.1

```{r}  
tt <- c(6, 7, 10, 15, 19, 25)
delta <- c(1, 0, 1, 1, 0, 1)
trt <- c(0, 0, 1, 0, 1, 1)

plsimple <- function(beta) {#log partial likelihood
  psi <- exp(beta)
  result <- log(psi) - log(3*psi + 3) - log(3*psi + 1) - log(2*psi + 1)
  result
  }

beta.vec <- ((-200):50) / 50
plikel.vec <- plsimple(beta.vec)

plot(plikel.vec ~ beta.vec, type="l", xlab=expression(beta), 
     ylab="log partial likelihood")
title('Log Partial Likelihood, Fig 5.1')

#Find MLE:  this method allows for constraints
result <- optim(par=0, fn = plsimple, method = "L-BFGS-B",   
                control=list(fnscale = -1), lower = -3, upper = 1)
result$par
abline(v=result$par, lty=2, col='blue')
abline(v=0, lty=2, col='red')
```




# Section 5.3  Three types of tests

```{r}
beta.hat <- result$par
beta.hat  #hazard ratio
exp(beta.hat)
plmax <- plsimple(beta.hat)
plmax

# the text changed variable names
status <- delta
grp <- trt

# Fit a Cox Prop Hazards model
result.cox <- coxph(Surv(tt, status) ~ grp)

summary(result.cox)
logLik(result.cox)
coef(result.cox)
```

## Interpret the results
Hazard of death for treated is `r round(exp(beta.hat), 3)` times that for control.
However, the confidence interval for the hazard ratio
is (`r round(summary(result.cox)$conf.int[,3:4], 3)`), 
which does include 1, indicating that this result is
not significant.



## Other outputs
Rsquare = reflects the improvement in the fit of the model with the covariate
compared with a model having no covariates (null model).  Bigger is better.

Concordance = C-Statistics = Harrell's C = measure of the predictive
discrimination of a covariate.  It lies between 0 and 1, and the bigger,
the better.


## Wald test

```{r}
#install.packages("numDeriv")    # must be done once
#library(numDeriv)   

beta.hess <- hessian(func=plsimple, x=result$par)
beta.hess
beta.se <- 1 / sqrt(-beta.hess)  # standard error
beta.se


#wald <- result$par*sqrt(-hessian(func=plsimple, x=result$par))
wald <- beta.hat / beta.se
wald   # Wald test
2 * pnorm(abs(wald), lower.tail=F)   #since only 1 parameter, use N(0,1)

pchisq(wald^2, 1, lower.tail=F)     #general case uses chi-square(#param)


sqrt(result.cox$wald.test)  # compare to model output
```


## Likelihood ratio test

```{r}

beta.lr <- 2 * (plsimple(beta.hat) - plsimple(0))   # likelihood ratio test
beta.lr
pchisq(beta.lr, 1, lower.tail=F)     #general case uses chi-square(#param)
```


## Score test

```{r}
beta.gr0 <- grad(func=plsimple, x=0)
beta.gr0
beta.hess0 <- hessian(func=plsimple, x=0)
beta.hess0
score.test <- beta.gr0^2 / (-beta.hess0)
score.test

pchisq(score.test, df=1, lower.tail=F) # score test


survdiff(Surv(tt, status) ~ grp)


```


# The pharmaco Smoking example in the library(asaur)
## Fit two Cox models and compare survival curves

```{r}
#?pharmacoSmoking
#library(asaur)
#library(survival)
summary(pharmacoSmoking)
phvars <- names(pharmacoSmoking)
phvars <- phvars[!is.element(phvars, c('id','grp'))]
phvars

#library(tableone)
# You will find a Table 1, similar to this, is most medical papers.
print(CreateTableOne(phvars, c('grp'), data=pharmacoSmoking),
         showAllLevels=TRUE)


pharmaco.cox1 <- coxph(Surv(ttr,relapse) ~ grp, data=pharmacoSmoking)
summary(pharmaco.cox1)
AIC(pharmaco.cox1)
L1 <- logLik(pharmaco.cox1)
L1
2*(-L1 + 1)  #same as AIC

ph.newdat1 <- data.frame(grp=levels(pharmacoSmoking$grp))
ph.newdat1
plot(survfit(pharmaco.cox1, newdata=ph.newdat1),
     lty=1:2, col=1:2, ylab='Survival Prob', xlab='days')
legend('topright', levels(pharmacoSmoking$grp), lty=1:2,
       col=1:2, bty='n')
title('Relapse-free survival, Pharmaco Smoking data')


pharmaco.cox2 <- coxph(Surv(ttr,relapse) ~ grp + ageGroup2 + gender + levelSmoking,
                       data=pharmacoSmoking)
summary(pharmaco.cox2)
AIC(pharmaco.cox2)
ph.newdat2 <- data.frame(grp=levels(pharmacoSmoking$grp),
                        ageGroup2='21-49', gender='Female', levelSmoking='heavy')
plot(survfit(pharmaco.cox2, newdata=ph.newdat2),
     lty=1:2, col=1:2, ylab='Survival Prob', xlab='days')
legend('topright', levels(pharmacoSmoking$grp), lty=1:2,
       col=1:2, bty='n')
title('Relapse-free survival, Female, 21-49, heavy')
L2 <- logLik(pharmaco.cox2)
L2

L2-L1
pchisq(2*(L2-L1), 3, lower.tail = FALSE)

anova(pharmaco.cox1, pharmaco.cox2)


```





# The pharmaco Smoking example using library(rms) and cph
## Compare survival curves for the pharmaco Smoking example

```{r}
#install.packages("rms")
#library(rms)
dd <- datadist(pharmacoSmoking)
options(datadist='dd')

pharmaco.cph1 <- cph(Surv(ttr,relapse) ~ grp, data=pharmacoSmoking,
                      x=TRUE, y=TRUE)
summary(pharmaco.cph1)
survplot(pharmaco.cph1)

pharmaco.cph2 <- cph(Surv(ttr,relapse) ~ grp + ageGroup2 + gender + levelSmoking,
                       data=pharmacoSmoking, x=TRUE, y=TRUE)
summary(pharmaco.cph2)
survplot(pharmaco.cph2, grp, col=1:2)
title('Relapse-free survival, Female, 21-49, heavy')

options(datadist=NULL)
```




