---
title: "Chapter 10"
author: "Mary Lesperance"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: 3
geometry: margin=.75in
fontsize:  11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=5, fig.width=6.5, warning=FALSE)
```

# Section 10.3.1   Assessing the Weibull distribution

```{r Weibull}
library(survival)
library(asaur)
#install.packages(c("asaur")) # must be done once
#install.packages("Hmisc")   # must be done once
library(Hmisc)

#Convert time in weeks to months
gastricXelox <- within( gastricXelox, timeMonths <-timeWeeks*7/30.25 )
summary(gastricXelox)

#KM curve
gas.km <- survfit(Surv(timeMonths, delta) ~ 1, data=gastricXelox)

logLogSurvEst <- log(-log(gas.km$surv)) #log-log(S(t))
logSurvTime <- log(gas.km$time)  #log(t)

# Fig 10.1 - cloglog SurvTime versus log SurvTime
plot(logLogSurvEst ~ logSurvTime) 
title('gastricXelox Cloglog Survival vs Log Time')

#Fit straight line through data
gas.lm <- lm(logLogSurvEst ~ logSurvTime)
gas.lm

# intercept estimates -mu/sigma; slope estimates 1/sigma
abline(gas.lm)
# the Weibull model is not appropriate

###-------------------
# Redo for pharmacoSmoking data
#attach(pharmacoSmoking) 
summary(pharmacoSmoking)
#amend one zero time
pharmacoSmoking <- within(pharmacoSmoking, ttr[ttr == 0]  <- 0.5)
summary(pharmacoSmoking)

pharma.km <- survfit(Surv(ttr, relapse) ~ 1, data=pharmacoSmoking) 

logLogSurvEst <- log(-log(pharma.km$surv))
logSurvTime <- log(pharma.km$time)
pharma.lm <- lm(logLogSurvEst ~ logSurvTime) 
pharma.lm
# intercept estimates -mu/sigma; slope estimates 1/sigma

pharma.lm$coef
sigma.est <- 1 / pharma.lm$coef[2]
mu.est <- -pharma.lm$coef[1] * sigma.est
mu.est; sigma.est

# Fig 10.2 - cloglog SurvTime versus log SurvTime
plot(logLogSurvEst ~ logSurvTime) 
abline(pharma.lm)
title('pharmacoSmoking Cloglog Survival vs Log Time')
# the Weibull model appears to be a good fit

```


# Section 10.3.2   Weibull maximum likelihood

```{r Weibull-max}
# see log likelihood function in notes
# par: c(mu, sigma), tt: time, status: event indicator
# In the (lambda, alpha) parameterization:  mu=-log(lambda), sigma=1/alpha
logLikWeib <- function(par, tt, status) {
   mu <- par[1]
   sigma <- par[2]
   lambda.p <- exp(-mu)
   alpha.p <- 1/sigma

   dd <- sum(status)
   sum.t <- sum(status*log(tt))
   sum.t.alpha <- sum(tt^alpha.p)

   term.1 <- dd*log(alpha.p) + alpha.p*dd*log(lambda.p)
   term.2 <- (alpha.p - 1)*sum.t
   term.3 <- (lambda.p^alpha.p)*sum.t.alpha
   result <- term.1 + term.2 - term.3
   result    
   }

#Find mu.hat, sigma.hat, the MLE's
#Use mu.est, sigma.est as starting values for MLE's
pharma.weib <- with(pharmacoSmoking, optim(par=c(mu.est, sigma.est), fn=logLikWeib, method="L-BFGS-B",
    lower=c(0.001, 0.01), upper=c(5, 5),
    control=list(fnscale = -1),
    tt=ttr, status=relapse))  
pharma.weib$par

#survreg does Weibull regression using log-linear model (AFT model)
result.survreg.0 <- survreg(Surv(ttr, relapse) ~ 1, dist="weibull", data=pharmacoSmoking)
summary(result.survreg.0)


```

# Section 10.3.3   Profile Weibull likelihood 

 
```{r Profile-Weibull}
#Maximize the log-likelihood over lambda, for a fixed value of alpha
# The Mle of lambda given alpha is:  (dd/sum.t.alpha)^(1/alpha.p)
# Substitute this into the log likelihood
# The result is called the Profile Log Likelihood

logLikWeibProf <- function(par, tt, status) {
   # find log-likelihood for a particular sigma, using mle for mu
   sigma <- par
   alpha.p <- 1/sigma
   dd <- sum(status)
   sum.t <- sum(status*log(tt))
   sum.t.alpha <- sum(tt^alpha.p)
   lambda.p <- (dd/sum.t.alpha)^(1/alpha.p)

   term.1 <- dd*log(alpha.p) + alpha.p*dd*log(lambda.p)
   term.2 <- (alpha.p - 1)*sum.t
   term.3 <- (lambda.p^alpha.p)*sum.t.alpha
   result <- term.1 + term.2 - term.3
   result 
   }
   
pharma.Prof <- with(pharmacoSmoking, optim(par=c(sigma.est), fn=logLikWeibProf, method="L-BFGS-B",
     lower=c(0.01), upper=c(5), control=list(fnscale = -1),
     tt=ttr, status=relapse) ) 
sigma.hat <- pharma.Prof$par 
sigma.hat  

#Compare estimate for mu
dd <- sum(pharmacoSmoking$relapse)
alpha.p <- 1/sigma.hat
sum.t.alpha <- sum(pharmacoSmoking$ttr^alpha.p)
lambda.p <- (dd/sum.t.alpha)^(1/alpha.p)
mu.hat <- -log(lambda.p)
mu.hat 

#Plot profile log-likelihood for a list of sigma values
sigma.list <- seq(1, 5, length=400)
n.list <- length(sigma.list)
logLik.list <- rep(NA, n.list)
for (i in 1:n.list) {
  logLik.list[i] <- with(pharmacoSmoking, logLikWeibProf(par=sigma.list[i], ttr, relapse))
  } 

# Fig. 10.3
plot(logLik.list ~ sigma.list, type="l", xlab="sigma", 
   ylab="profile log-likelihood") 
abline(v=sigma.hat, col="gray")
title('pharmacoSmoking Profile Log-likelihood for Weibull sigma')
```



# Section 10.3.4  Selecting a Weibull distribution - OMIT THIS SECTION 
   

```{r figSelectWeib}
# install.packages("Hmisc")   # must be done once
# library(Hmisc)

#KM curve
pharma.surv <- survfit(Surv(ttr, relapse) ~ 1, subset={grp =="patchOnly"},
                       data=pharmacoSmoking)
pharma.survsum <- summary(pharma.surv, time=c(28, 84))

t.vec=pharma.survsum$time 
s.vec=pharma.survsum$surv
cbind(t.vec, s.vec)

#Match 2 points from KM curve with Weibull model
pharmWeib <- Weibull2(t.vec, s.vec)  #Weibull survival function
pharmWeib

#Compare distributions, MLE, 2 point and KM

#2 point match
t.vals <- 1:200
s.vals <- pharmWeib(t.vals)

#MLE, reparameterize
pharm.weib.basic <- survreg(Surv(ttr, relapse) ~ 1,
     dist="weibull", subset={grp =="patchOnly"}, data=pharmacoSmoking )
mu.hat <- pharm.weib.basic$coefficients
sigma.hat <- pharm.weib.basic$scale
lambda.hat <- exp(-mu.hat)      # " 1 / scale"
alpha.hat <- 1/sigma.hat        # "shape"
s.mle.vals <- 1 - pweibull(t.vals, shape=alpha.hat, scale=1/lambda.hat)

# Fig. 10.4

plot(pharma.surv, conf.int=F, lwd=2, xlab="Days to relapse",
   ylab="Survival probability", cex.lab=1.3, cex.axis=1.3)
lines(s.mle.vals ~ t.vals, col="blue", lwd=2)
lines(s.vals ~ t.vals, col="red", lwd=2)
points(t.vec, s.vec, col="red", pch=16, cex=1.5)
legend("topright", legend=c("Maximum likelihood", "Match at 24 and 84 days"),
  col=c("blue", "red"), lwd=2)


```


# Section 10.3.6  Compare accelerated failure times and proportional hazards

```{r AccelFailure}
#Compare models with one covariate grp
# Weibull model
pharma.survreg.grp <- survreg(Surv(ttr, relapse) ~ grp, dist="weibull", 
                              data=pharmacoSmoking)
summary(pharma.survreg.grp)

#Plot estimated survival curves using both methods
#Weibull convert estimates in (mu, sigma) parameterization to (lambda, alpha)
mu0.hat <- pharma.survreg.grp$coef[1]
sigma.hat <- pharma.survreg.grp$scale 
alpha.hat <- 1/sigma.hat 
lambda0.hat <- exp(-mu0.hat)
tt.vec <- 0:182 #time vector

#Weibull estimate of S0(t) for the combination group
surv0.vec <- 1 - pweibull(tt.vec, shape=alpha.hat, scale=1/lambda0.hat)

#Weibull estimate of S(t;z=1) for the patchOnly group
gamma.hat <- pharma.survreg.grp$coef[2]
surv1.vec <- surv0.vec^(exp(-gamma.hat/sigma.hat))

#Cox model survival curve estimate for combination/patch
#Cox model
pharma.coxph.grp <- coxph(Surv(ttr, relapse) ~ grp, data=pharmacoSmoking)
summary(pharma.coxph.grp) 

pharma.coxph.Sest <- survfit(pharma.coxph.grp, 
    newdata=data.frame(list(grp=c("combination","patchOnly"))))
    
# Fig. 10.5
#The Weibull and Cox model estimates are similar
plot(pharma.coxph.Sest, col=c("red", "black"), xlab="Time in days",
  ylab="Survival probability", cex.axis=1.5, cex.lab=1.5, lwd=2)
lines(surv0.vec ~ tt.vec, col="red", lty=2)
lines(surv1.vec ~ tt.vec, col="black", lty=2)
legend("topright", legend=c("Cox model combination", "Cox model patch",
       "Weibull model combination", "Weibull model  patch"),
       col=c("red", "black", "red", "black"), lty=c(1,1,2,2))

```



# Section 10.3.7  Weibull model with multiple covariates


```{r Weibullcov}
#Weibull model
pharma.weib <- survreg(Surv(ttr, relapse) ~ grp + age + employment, dist="weibull",
                      data=pharmacoSmoking)
# Negative coefficients imply shorter lifetimes
summary(pharma.weib)

weib.coef.all <- pharma.weib$coef 
weib.coef <- weib.coef.all[2:5]
weib.coef.ph <- -weib.coef/pharma.weib$scale

#Cox model
pharma.coxph <- coxph(Surv(ttr, relapse) ~ grp + age + employment,
                     data=pharmacoSmoking )
summary(pharma.coxph)
coxph.coef <- pharma.coxph$coef 

#The two models provide similar estimates
data.frame(weib.coef.ph, coxph.coef)
```

# Section 10.3.8  Weibull model selection and residual analysis
```{r Weibdiag}
#Fit full weibull model

pharma.All.weib <- survreg(Surv(ttr, relapse) ~ grp + gender + 
                   race + employment + yearsSmoking + levelSmoking +
                   age + priorAttempts + longestNoSmoke, dist="weibull",
                   data=pharmacoSmoking) 
summary(pharma.All.weib)

#Weibull backward model
pharma.step.weib <- step(pharma.All.weib, direction="backward")
summary(pharma.step.weib)

pharma.devres <- residuals(pharma.step.weib, type="deviance")

# the following function is from the appendix:
smoothSEcurve <- function(yy, xx) {
  # use after a call to "plot"
  # fit a lowess curve and 95% confidence interval curve
  xx.list <- min(xx) + ((0:100)/100)*(max(xx) - min(xx)) # make list of x values

  # Then fit loess function through the points (xx, yy) at the listed values
  yy.xx <- predict(loess(yy ~ xx), se=T, newdata=data.frame(xx=xx.list))

  lines(yy.xx$fit ~ xx.list, lwd=2)
  lines(yy.xx$fit - qt(0.975, yy.xx$df)*yy.xx$se.fit ~ xx.list, lty=2)
  lines(yy.xx$fit + qt(0.975, yy.xx$df)*yy.xx$se.fit ~ xx.list, lty=2)
  }

# Fig. 10.6
oldpar <- par(mfrow=c(2,2)) 
plot(pharma.devres  ~ age, data=pharmacoSmoking) 
smoothSEcurve(pharma.devres, pharmacoSmoking$age) 
title("Deviance residuals\nversus age")

plot(pharma.devres  ~ grp, data=pharmacoSmoking)   
title("Deviance residuals\nversus treatment group")

plot(pharma.devres  ~ employment, data=pharmacoSmoking) 
title("Deviance residuals\nversus employment") 


pharma.dfbetas <- residuals(pharma.step.weib, type="dfbetas")
summary(pharma.dfbetas)
dimnames(pharma.dfbetas)[[2]] <- c(names(coef(pharma.step.weib)), 'Log(scale)')

n.dfbetas <- nrow(pharma.dfbetas)
index.obs <- 1:n.dfbetas 

# Fig. 10.7
summary(pharma.dfbetas)
plot(pharma.dfbetas[,5] ~ index.obs, type="h",
   xlab="Observation", ylab="Change in coefficient")
title('dfbetas for Age')
abline(h=0)

#identify(resid.dfbeta[,3] ~ index.obs)   # use mouse to click points to identify
#  press "Esc" key to exit
par(oldpar)

```

# Section 10.4 Other parametric distributions

```{r}
#Negative coefficients imply shorter lifetimes
model.pharm.lognormal <- survreg(Surv(ttr, relapse) ~ grp + age +
    employment, dist="lognormal", data=pharmacoSmoking)
summary(model.pharm.lognormal)

model.pharm.loglogistic <- survreg(Surv(ttr, relapse) ~ grp + 
    age + employment, dist="loglogistic", data=pharmacoSmoking)
summary(model.pharm.loglogistic)

```

