---
title: "Chapter 9"
author: "Mary Lesperance"
date: '`r Sys.Date()`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: 3
geometry: margin=.75in
fontsize:  11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=5, fig.width=6.5)
```

# Section 9.2.1 competing risks   # prostateSurvival is in "asaur" package

```{r Compete}
library(survival)
library(asaur)
#install.packages(c("mstate", "cmprsk")) # must be done once
library(mstate)
library(cmprsk)

summary(prostateSurvival)

#Two outcomes
#status =1 (death due to prostate cancer), =2 (death due to other cause)
prostateSurvival <- within(prostateSurvival, {
   status.prost <- as.numeric({status == 1}) 
   status.other <- as.numeric({status == 2})}) 
ls()   #these variables are not in the workspace
summary(prostateSurvival)

#attach(prostateSurvival)

#Choose high risk patients
prostateSurvival.highrisk <- subset(prostateSurvival, {{grade == "poor"} & 
   {stage=="T2"} & {ageGroup == "80+"}})
head(prostateSurvival.highrisk) 

#Compare survival prob estimates for 2 causes of death
result.prostate.km <- survfit(Surv(survTime, event=status.prost) ~ 1,
    data=prostateSurvival.highrisk)
result.other.km <- survfit(Surv(survTime, event=status.other) ~ 1,
    data=prostateSurvival.highrisk) 

attributes(result.prostate.km)

surv.other.km <- result.other.km$surv #S(t)
cumDist.other.km <- 1 - surv.other.km   #F(t)=1-S(t)
surv.prost.km <- result.prostate.km$surv  #S(t)
cumDist.prost.km <- 1- surv.prost.km  #F(t)=1-S(t)
time.km <- result.other.km$time / 12   #time is in months

#Plot F(t) cdf estimates versus time
plot(cumDist.prost.km ~ time.km, type="s", ylim=c(0,1),  
    xlab="Years from prostate cancer diagnosis",  
    ylab='probabilities', col=1, lty=1) 
lines(cumDist.other.km ~ time.km, type="s", col=2, lty=2) 
legend('topleft',c('Prostate Death, 1-km','Other Death, 1-km'),
       col=1:2, lty=1:2, bty='n')
title('Figure 9.1b')

#summary(result.prostate.km,c(119))  #almost 10-year survival probability

#F(t) Probability of events by 119 months
1 - summary(result.prostate.km,c(119))$surv   #almost 10-year probabilities
1 - summary(result.other.km,c(119))$surv

#These are mutually exclusive events, but(!) probs sum to >1
```

```{r}
# Figure 9.1  - see text for description
plot(cumDist.prost.km ~ time.km, type="s", ylim=c(0,1),  
    xlab="Years from prostate cancer diagnosis",  
    ylab='probabilities', col=1, lty=1) 
lines(surv.other.km ~ time.km, type="s", col=2, lty=2) 
legend('topright',c('Prostate Death, 1-km','Other Death, km'),
       col=1:2, lty=1:2, bty='n')
title('Figure 9.1')

#detach(prostateSurvival)
```

# Section 9.2.2 cause-specific hazards and cumulative incidence functions

```{r cause-specific}
#Simple example
tt <- c(2,7,5,3,4,6)
status <- c(1,2,1,2,0,0) #two causes of death, 1 and 2
status.any <- as.numeric(status >= 1) 
result.any <- survfit(Surv(tt, status.any) ~ 1) #overall survival function
summary(result.any) 
result.any$surv


#install.packages("mstate") # must be done once
#library(mstate)

#Cuminc computes cumulative incidence

simple.ci <- Cuminc(time=tt, status=status, failcodes=1:2)
simple.ci



## Aside FYI
#create same table using survfit adn type="mstate"
simple.survmstate  <- survfit(Surv(tt, status, type="mstate") ~ 1)
simple.survmstate.sum <- summary(simple.survmstate)
simple.survmstate.sum
attributes(simple.survmstate.sum)

# bind table
simple.ci2 <- cbind(time=simple.survmstate.sum$time, n.risk=simple.survmstate.sum$n.risk[,3],
            simple.survmstate.sum$pstate, simple.survmstate.sum$std.err)
colnames(simple.ci2)[-(1:2)] <- c('CI.1','CI.2','Surv','seCI.1','seCI.2','seSurv')
simple.ci2


```



## Figure 9.4  

```{r fig9_4}
# Prostate cancer data competing risks analysis compared with km

summary(prostateSurvival.highrisk)

#frequencies for each competing risk
summary(factor(prostateSurvival.highrisk$status))

# Cuminc results using survfit and type="mstate" 
pros.sf <- summary(survfit(Surv(survTime, status, type="mstate") ~ 1, 
                      data=prostateSurvival.highrisk))

attributes(pros.sf)
pros.times <- pros.sf$time/12
pros.CIs <- pros.sf$pstate  

pros.ci1 <- pros.CIs[,1]  #ci for prostate
pros.ci2 <- pros.CIs[,2]  #ci for other

pros.Cci2 <- 1 - pros.ci2  #complement of ci2

#This graph is in textbook
plot(pros.Cci2 ~ pros.times, type="s", ylim=c(0,1), lwd=2, col=1, lty=1,
    xlab="Time in years", ylab="Survival probability",
    xlim=c(0,10))
lines(pros.ci1 ~ pros.times, type="s", lwd=2, col=2, lty=1) 
lines(surv.other.km ~ time.km, type="s", col=1, lwd=1, lty=2)
lines(cumDist.prost.km ~ time.km, type="s", col=2, lwd=1, lty=2) 
legend('topright', c('1-ci other','ci prostate','km other','1-km prostate'),
       col=c(1,2,1,2), lty=c(1,1,2,2), lwd=c(2,2,1,1), bty='n')
title('Comparison cumulative incidence and KM')

```



## Figure 9.5 - these are not stacked curves as in the book  

We compare the cumulative incidence curves for the two competing risks.
This graph is not in textbook.

```{r fig9_5}
plot(survfit(Surv(survTime, status, type="mstate") ~ 1,  
             data=prostateSurvival.highrisk), lty=1:2, col=1:2,  ylim=c(0,1),
             xscale=12, xlab="Time in years", ylab="Survival probability")         
            
title('Cumulative incidence functions')    
legend('topleft', c('Prostate cancer', 'Other causes'), lty=1:2, col=1:2, bty='n' )

```


# Section 9.2.4 regression methods for cause-specific hazards (Fine and Gray)

I recommend crr method for regression modelling of competing risks.

```{r crr_reg}
prostateSurvival.T2 <- prostateSurvival[prostateSurvival$stage=="T2",]
summary(prostateSurvival.T2)

#attach(prostateSurvival.T2)

#Putter etc. simply treat competing risk as censored
prostate.cox <- coxph(Surv(survTime, status.prost) ~ grade + ageGroup,
                         data=prostateSurvival.T2)
summary(prostate.cox) 

other.cox <- coxph(Surv(survTime, status.other) ~ grade + ageGroup,
                       data=prostateSurvival.T2)
summary(other.cox) 

#install.packages("cmprsk") # must be done once
#library(cmprsk)
#proportional subdistribution hazards regression model described in Fine and Gray (1999)

#cmprsk package does not support formula - you need to supply covariate matrix
cov.matrix <- model.matrix(~ grade + ageGroup, data=prostateSurvival.T2)
head(cov.matrix)

#crr does not have the data option, use with()
#Compare covariate effects on sub-hazards for the two competing risks.
result.prostate.crr <-with(prostateSurvival.T2,  crr(survTime, status, 
                  cov1=cov.matrix[,-1], failcode=1))
summary(result.prostate.crr)

result.other.crr <- with(prostateSurvival.T2, crr(survTime, status, 
                  cov1=cov.matrix[,-1], failcode=2) )
summary(result.other.crr)

```



# Section 9.2.5 Covariates on different causes of death

Here a several models that use multi-state set-up for the analysis of
competing risks.  See textbook for discussion.

```{r Covar_Cause}
library(mstate)
tmat <- trans.comprisk(2, names = c("event-free", "prostate", "other"))
tmat

prostate.long <- with(prostateSurvival.T2, msprep(time = cbind(NA, survTime, survTime), 
      status = cbind(NA, status.prost, status.other),  
      keep = data.frame(grade, ageGroup), trans = tmat))

head(prostate.long) ; head(prostateSurvival.T2)
events(prostate.long) 

summary(coxph(Surv(time, status) ~ grade + ageGroup, data=prostate.long, 
              subset={trans==1}))
summary(coxph(Surv(time, status) ~ grade + ageGroup, data=prostate.long, 
              subset={trans==2}))

#model assumes grade acts same on prostate and other causes
summary(coxph(Surv(time, status) ~ grade + ageGroup + strata(trans), data=prostate.long))

#grade acts differently
summary(coxph(Surv(time, status) ~ grade*factor(trans) + ageGroup + strata(trans), data=prostate.long))  

#grade and age act differently
summary(coxph(Surv(time, status) ~ (grade + ageGroup)*trans + ageGroup + strata(trans), data=prostate.long)) 


```
