---
title: "Chapter 3 - Nonparametric Survival Curve Estimation"
author: "Mary Lesperance"
date: "September 17, 2018"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=5, fig.width=6.5)
```

# Nonparametric estimation of the Survival Curve

```{r KMcurve}
library(survival)
tt <- c(7,6,6,5,2,4)
cens <- c(0,1,0,0,1,1)
Surv(tt, cens)
result.km <- survfit(Surv(tt, cens) ~ 1, conf.type="log-log")
result.km  #typo in book Table 3.1
summary(result.km)
plot(result.km)
title('Kaplan-Meier curve example')

abline(h=.5, col="red")

result.linear.km<- survfit(Surv(tt, cens) ~ 1, conf.type="plain")
result.linear.km  
summary(result.linear.km  )

```

# Nelson-Aalen estimate (Fleming-Harrington)

```{r}
result.fh <- survfit(Surv(tt, cens) ~ 1, conf.type="log-log",
            type="fh")
summary(result.fh)

```

# Example 1.2: gastricXelox

We now consider data from an actual clinical trial. The data set 'gastricXelox'
is a Phase II (single sample) clinical trial of the chemotherapeutic agent Xelox
administered to patients with advanced gastric cancer prior to surgery (Wang
et al. [74]). The primary outcome of interest is 'progression-free survival.' This
quantity is defined as the time from entry into a clinical trial until progression or
death, whichever comes first. The survival data set was extracted from the paper,
and the survival times rounded to the nearest week. The product-limit estimator
may be estimated and plotted as follows, after converting the time scale from weeks
to months.

```{r gastric}
library(asaur)
timeMonths <- gastricXelox$timeWeeks*7/30.25   #I would use 365.25/12=30.4375 instead of 30.25
delta <- gastricXelox$delta
gastric.km <- survfit(Surv(timeMonths, delta) ~ 1,
      conf.type="log-log")
plot(gastric.km, conf.int=T, mark="|", xlab="Time in months",
   ylab="Survival probability")
title("Progression-free Survival in Gastric Cancer Patients")
abline(h=.5,col=2)



```


# Median survival and confidence interval

```{r KMmedian}
gastric.km

summary(gastric.km, c(12, 24, 36))
```

# Median follow-up time

One measure of the quality of a clinical trial is the duration of follow-up, as
measured by the median follow-up time. This is a measure that captures how
long, on average, patients have been followed.

A simple definition is to consider all of the survival times, whether
censored or not, and find the median. A disadvantage of this is that a trial with
many early deaths, but a long observation period, would appear not to have a long
median follow-up time. A perhaps better way of looking at median survival is the
'potential' median survival. To obtain this estimate, one first switches the censoring
and death indicators, so that a “censored” observation is the “event”, while a death
is viewed as a censored observation, in the sense that the observation time would
have been much longer had the patient not died. One then computes the Kaplan-
Meier “survival” estimate using these reversed censoring indicators, and finds the
median survival, as discussed in the previous section. This method is also known as
the “reverse” Kaplan-Meier [59].
Most clinicians with whom I have worked ask for both meaures of follow-up time.

```{r}
delta.followup <- 1 - delta
survfit(Surv(timeMonths, delta.followup) ~ 1)

median(timeMonths)
```

# Hazard estimation

Estimates of the cumulative hazard can be used for two purposes:
1.  To assist with the selection of an appropriate 
parametric model;
2.  Crude estimates of the hazard rates can be obtained from
the slope of the cumulative hazard function.

```{r}
plot(gastric.km, conf.int=T, mark="|", xlab="Time in months",
   fun="cumhaz", ylab="Cumulative hazard" )
title("Cumulative hazard in Gastric Cancer Patients")

gastric.fh <- survfit(Surv(timeMonths, delta) ~ 1,   #Fleming-Harrington
      conf.type="log-log", type='fh')
plot(gastric.fh, conf.int=T, mark="|", xlab="Time in months",
   fun="cumhaz",ylab="Cumulative hazard")
title("Cumulative hazard (fh) in Gastric Cancer Patients")

# plot the two types on one graph
plot(gastric.km, conf.int=F, mark="|", xlab="Time in months",
   fun="cumhaz", ylab="Cumulative hazard" )
lines(gastric.fh, conf.int=F, mark="|", xlab="Time in months",
   fun="cumhaz",ylab="Cumulative hazard", col=2, lty=2)
legend("bottomright",c("KM","fm"),col=1:2,lty=1:2)
title("Two types of Cumulative hazards, Gastric Cancer patients")

```

